Constraint Grammar as a SAT problem

### 2015-06-01 paper/nodalida2015.tex
add abstract

%\begin{abstract} \end{abstract}

\section{Introduction}

We represent Constraint Grammar (CG) \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
This is attractive from several reasons: formal logic is
well-studied, and serves as an abstract language to reason about the
properties of CG. Constraint rules encoded in logic capture richer
dependencies between the tags than standard CG.

@@

Encoding CG in logic increases computational complexity, 
but as a tradeoff, it could simplify the grammar writing.
Due to representing the rules as implications, cautious context
condition becomes irrelevant.
One rule corresponds to many logical formulas (each condition being
the target), which makes the number of the rules potentially smaller.
Ordering can be preserved or discarded; in the latter case, we solve
eventual rule conflicts by finding a solution that discards the least
number of rule applications.
% our experiments show that it is possible to write a small grammar, 
% which, when run with SAT-CG, disambiguates more than the same
%grammar run with VISL CG-3.

We test our implementation by parsing texts in the order of
10,000s--100,000s words, using grammars with hundreds of rules.

%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.

\section{Related work}
\label{sect:related}

Our work is inspired by \cite{lager98}, which presents constraint rules as a disjunctive logic program,
and \cite{lager_nivre01}, which reconstructs 4 different formalisms from a logical point of view.
\cite{lindberg_eineborg98ilp,asfrent14} use Inductive Logic Programming to learn CG rules from a tagged corpus.
We explore different approaches to ordering. In the case of unordered
rules, it resembles Finite-State Intersection Grammar \cite{koskenniemi90}.
See also \cite{lager01transformation} for discussion of ordering of the CG rules.
\section{CG as a SAT problem}
Let us demonstrate our approach with the following example in Spanish.

@@

\section{Results and evaluation}
\label{sec:eval}

@@

We tested the performance by parsing Don Quijote (384,155 words) with
the same Spanish grammars as in the previous experiment. 
Table~\ref{table:time} shows execution times compared to VISL CG-3;
SAT-CG\textsubscript{U} is the unordered scheme and
SAT-CG\textsubscript{O} is the ordered.
% For both systems, the number of rules in the grammar affects the performance more than the word count.
From the SAT solving side, maximisation is the most costly operation. 
Emulating order is slightly faster, likely because the maximisation problems are smaller.
In any case, SAT does not seem to be the bottleneck: with 261 rules,
the maximisation function was called 147,253 times, and with 19 rules,
132,255 times, but 
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
the differences in the execution times are much larger, which suggests
that there are other reasons for the worse performance. 
This is to be expected, SAT-CG is currently just a naive
proof-of-concept implementation with no optimisations.

@@

\section{Applications and Future Work}
\label{sec:apps}
As described in Section~\ref{sec:eval}, our implementation is not
competetive with the state of the art. It is more fruitful to focus on
things where the SAT-based representation can make an impact.
One such thing is grammar analysis\footnote{We thank Eckhard Bick for
  the suggestion.}.
There has been work on automatic tuning of hand-written CGs
\cite{bick2013tuning}, but to our knowledge, no tools which would
search for inconsistencies or unoptimal design in order to point it out to the grammar writer.

The sequential application of traditional CG rules is good for
performance and transparency. As soon as a rule applies, the analyses
are removed from the input, and next rules get the modified sentence
as their input. 
% At the execution of rule, there is no way to go back to earlier rules and undo them. 
Thus there is no way to know what part of the input comes from the
morphological analyser and what part from applying previous rules.

>>>>

\begin{abstract}

We represent Constraint Grammar (CG) as a Boolean satisfiability (SAT) problem.
Encoding CG in logic increases the computational complexity, 
but brings some new features to the grammars.
The rules are interpreted in a more declarative way, which makes
cautious context and ordering irrelevant.
One rule corresponds to many logical formulas,
where each condition interpreted as a target---this makes the number of the rules
potentially smaller. 
Ordering can be preserved or discarded;
in the latter case, we solve eventual rule conflicts by finding a
solution that discards the least number of rule applications.
We test our implementation by parsing texts in the order of
10,000s--100,000s words, using grammars with hundreds of rules.
\end{abstract}

\section{Introduction and previous research}

\todo{make this nicer: Constraint Grammar (CG)
  \cite{karlsson1995constraint} is a relatively young formalism.
In this work, we implement CG using a SAT solver,
thus relating CG to a wider context in the theory of computation.
This is attractive from several reasons: formal logic is
well-studied, and serves as an abstract language to reason about the
properties of CG. Constraint rules encoded in logic capture richer
dependencies between the tags than standard CG.}

@@

\todo{Our work is inspired by \cite{lager98}, which presents constraint rules as a disjunctive logic program,
and \cite{lager_nivre01}, which reconstructs 4 different formalisms from a logical point of view.
\cite{lindberg_eineborg98ilp,asfrent14} use Inductive Logic Programming to learn CG rules from a tagged corpus.
We explore different approaches to ordering. In the case of unordered
rules, it resembles Finite-State Intersection Grammar \cite{koskenniemi90}.
See also \cite{lager01transformation} for discussion of ordering of
the CG rules.}

\section{CG as a SAT problem}
Let us demonstrate our approach with the following example in Spanish.

@@

\section{Evaluation}
\label{sec:eval}

@@

We tested the performance by parsing Don Quijote (384,155 words) with
the same Spanish grammars as in the previous experiment. 
Table~\ref{table:time} shows execution times compared to VISL CG-3;
SAT-CG\textsubscript{u} is the unordered scheme and
SAT-CG\textsubscript{o} is the ordered.
% For both systems, the number of rules in the grammar affects the performance more than the word count.
From the SAT solving side, maximisation is the most costly operation. 
Emulating order is slightly faster, likely because the maximisation problems are smaller.
In any case, SAT does not seem to be the bottleneck: with 261 rules,
the maximisation function was called 147,253 times, and with 19 rules,
132,255 times, but 
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
the differences in the execution times are much larger, which suggests
that there are other reasons for the worse performance. 
This is to be expected, SAT-CG is currently just a naive
proof-of-concept implementation with no optimisations.

@@

\section{Applications}
\label{sec:apps}
As described in Section~\ref{sec:eval}, our implementation is not
competetive with the state of the art. It is more fruitful to focus on
things where the SAT-based representation can make an impact.
One such thing is grammar analysis\footnote{We thank Eckhard Bick for
  the suggestion.}.
There has been work on automatic tuning of hand-written CGs
\cite{bick2013tuning}, but to our knowledge, no tools which would
search for inconsistencies or unoptimal design in order to point it out to the grammar writer.

The sequential application of traditional CG rules is good for
performance and transparency. As soon as a rule takes action, the analyses
are removed from the input, and next rules get the modified sentence
as their input. 
% At the execution of rule, there is no way to go back to earlier rules and undo them. 
Thus there is no way to know which part comes directly from the
raw input and which part from applying previous rules.
